---
title: "Toronto shooting occurence"
author: 
  - SangWoong Lee
thanks: "Code and data are available at: LINK. https://github.com/SANGWOONG-LEE/starter_folder-main."
date: "06 Feburary 2022"
abstract: "Shooting occurence is important in the sense of public safety and protection.  We obtain and analyze yearly shooting occurence and region based in Toronto. From analysis, we obtain that shooting has increased over years as well as there exists some specific divisions where shooting occurs more often than others. Therfore, our findings have implications for more dispositions of police and related jobs into the divisions as well as concerns regarding the divisions"
output:
  bookdown::pdf_document2
toc: FALSE
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(palmerpenguins)
```

# Introduction

When I have conversations with people who are unfamiliar with N.American culture
and life style, the most questions they ask to me is whether Canada is safe from
shooting or not. I assume that this is because the U.S legalize the right of
the people to keep and bear arms. Since I am not living in the U.S, I am not 
100% certain regarding the situation of arms there. However, as far as I know, 
shooting in Canada especially Toronto is relatively safe from the U.S. Therefore, 
I decide to conduct analysis of shooting in Toronto. Then, I ask myself that 
what meanings the analysis will have other than just showing shootings frequency.
I figure out that if we analyze with respect to districts, it will be meaningful for people who are engaged in preventing such activity.
I conduct simple analysis of frequency of shooting in accordance with division and year. 
The remainder of this paper is Section \@ref(data) explains the data. Section \@ref(references) will show the reference. 

# Data
We obtain data from 'opendatatoronto', City of Toronto open Data portal.[@citeopendatatoronto] and the statistical programming language R [@citeR]. The data is composed of OccurredYear, GeoDivision and Count_. To be specifically, OccuredYear represents the year when shooting occurs. GeoDivison is the division where shooting occurs. Lastly, Count_ is how many times shooting occurs with respect to division and year. There exists limits to count parts because any veiled events are hard to identify. Other than that, it includes all general shooting occurrences such as firearm discharge, shooting event and person injured. Data is collected by police so it is very trustworthy, on the other hand, it might lack the actual feelings that citizens have. Also, we do not exactly know divisions that police is using, thus it is difficult to find that why specific divisions have more incidents than others. However, our purpose for this project is to find whether shooting occurrences increase over years or not, and what divisions have more incidents than others. Therefore, those limits do not violate much. By using, tidyverse package [@citetidyverse] and knitr package [@citeknitr], we could obtain desired table to explain data. 


```{r dataextract, echo = FALSE, warning = FALSE, message = FALSE}
library(tidyverse)
library(knitr)
shooting_count <- read_csv(here::here("inputs/data/shooting_count.csv"))

shooting_count |>
  select(OccurredYear, GeoDivision, Count_) |>
  slice(1:10) |> 
  kable(
    caption = "First ten rows of a dataset that shows shooting occurence",
    col.names = c("Occured year", "Division", "Occurence"),
    digits = 1,
    booktabs = TRUE,
    linesep = ""
  )
```

Table 1 below shows  10 rows of my data with respect to year, division and occurrence. It specifically shows the year 2014 according to division. One thing that we can first notice is that there exists wide range of the occurence in different divisions. For example, D31 has 31 reports of shootings, on the other hand division 11 has only 2 shooting occurrence. Later, I am going to conduct bar graph in order to find whether it is true in general. 

For this project, we are interested in the relationship between the division and shootings occurrence as well as year and shootings occurrence. By using ggplot[@citeggplot2], we can obtain clear graph of those.


```{r division, fig.cap="division", echo = FALSE, fig.height=3}
shooting_count |>  
  ggplot(mapping = aes(x = GeoDivision, y = Count_, fill = OccurredYear)) +
  geom_bar(stat='identity') + 
  labs(x = "Division", y = "Occurrence", fill = "Year", title = "Number of shooting occurences vs Divsion with respect to year")
```

Figure \@ref(fig:division) shows the distributions of year and division with respect to shooting occurrences. Clearly, as time goes, number of occurrences increase and as we have already seen in table \@ref(tab:dataextract), some divisions such as D31 and D23 have more reported cases than others. 
However, from figure \@ref(fig:division), it is bit ambiguous to observe that whether shooting occurrences increase or decrease. Therefore, I conduct separate graph regarding it. 

```{r year, fig.cap = "year", echo = FALSE,fig.height=3}
shooting_count |>  
  ggplot(mapping = aes(x = OccurredYear, y = Count_)) +
  geom_bar(stat = 'identity') + 
  labs(x = "Year", y = "Occurrence", title = "Yearly Shooting Occurences")
```

If you take a look at figure \@ref(fig:year) below, it is obvious to observe that the incidents are increasing over years. The last thing to notice that is change of shooting occurrence over years. As you have already seen, shooting occurrence increases over times. However, there exists some divisions with decreasing patterns such as D14, D31, D41, and D51. Implication for that is whatever efforts that police or related jobs put into those divisions are working. Last figure below shows the change of occurrences in each divisions. 

```{r change, fig.cap="change", echo = FALSE, fig.height = 4.5}
shooting_count |>  
    ggplot(aes(x=OccurredYear, y=Count_, group=GeoDivision, color=GeoDivision)) +
    geom_line() +
    labs(x = "Year", y = "Occurrence", title = "Change in shooting occurences with respect to Divisions")
```
It is a bit difficult to observe that which division corresponds to line since there are many lines. Let's put that besides, what we want from this project is observing decreasing patterns so we can learn successful strategies we have done in that specific region and year. Clearly from the graph, there seems many things that we can learn from what we have done in the corresponding divisions.


# References
This section shows all references I have used throughout this project. Also, I also need one extra reference to dplyr [@citedplyr].


